Reanalysis Step 5: Outcome Analysis with Three-Way Interaction

### 2025-09-01 reanalysis-5-outcome_analysis-interaction.qmd
consistecy on ix

```{r emmeans-analysis}
# Function to pool emmeans results across imputations
pool_emmeans_contrasts <- function(models, n_imputations) {

  emm_options(rg.limit = 500000)

  all_contrasts <- list()

@@

    emm <- emmeans(models[[i]], 
                   ~ ever_lapse_binary | dr_severity * any_treatment,
                   weights = "proportional")

    lapse_contrasts <- contrast(emm, method = "revpairwise", 
                               by = c("dr_severity", "any_treatment"))

    all_contrasts[[i]] <- as.data.frame(lapse_contrasts)
  }

  first_df <- all_contrasts[[1]]
  n_contrasts <- nrow(first_df)

  pooled_contrasts <- data.frame(
    dr_severity = first_df$dr_severity,
    any_treatment = first_df$any_treatment,
    contrast = first_df$contrast,
    estimate = numeric(n_contrasts),
    SE = numeric(n_contrasts),
    df = numeric(n_contrasts),
    stringsAsFactors = FALSE
  )

  # Pool each contrast using Rubin's rules
  for (j in 1:n_contrasts) {
    # Extract estimates and SEs for this contrast across imputations
    estimates <- sapply(all_contrasts, function(x) x$estimate[j])
    SEs <- sapply(all_contrasts, function(x) x$SE[j])

    # Apply Rubin's rules
    Q_bar <- mean(estimates)
    U_bar <- mean(SEs^2)  # Average of variances
    B <- var(estimates)   # Between-imputation variance
    T <- U_bar + (1 + 1/n_imputations) * B

    # Degrees of freedom
    r <- (1 + 1/n_imputations) * B / U_bar
    df_mi <- (n_imputations - 1) / (r^2)

    pooled_contrasts$estimate[j] <- Q_bar
    pooled_contrasts$SE[j] <- sqrt(T)
    pooled_contrasts$df[j] <- df_mi
  }

  return(pooled_contrasts)
}

# Pool contrasts across all imputations
pooled_lapse_effects <- pool_emmeans_contrasts(models, n_imputations)

# Calculate additional statistics
lapse_effects <- pooled_lapse_effects |>
  mutate(
    t_stat = estimate / SE,
    p.value = 2 * pt(-abs(t_stat), df = df),
    CI_Lower = estimate - qt(0.975, df = df) * SE,
    CI_Upper = estimate + qt(0.975, df = df) * SE
  )

# Create clean results table using lapse_effects which has all the calculated columns
subgroup_effects <- data.frame(
  DR_Severity = lapse_effects$dr_severity,
  Treatment = lapse_effects$any_treatment,
  Log_Odds = round(lapse_effects$estimate, 4),
  SE = round(lapse_effects$SE, 4),
  CI_Lower = round(lapse_effects$CI_Lower, 4),
  CI_Upper = round(lapse_effects$CI_Upper, 4),
  OR = round(exp(lapse_effects$estimate), 4),
  P_Value = format.pval(lapse_effects$p.value, digits = 4, eps = 0.0001)
)

kable(subgroup_effects,
      caption = "Effect of Lapsing Within Each DR Severity × Treatment Subgroup",
      row.names = FALSE)

# Store for visualization
subgroup_effects_stored <- subgroup_effects
```

@@

>>>>

```{r emmeans-analysis}
# Function to pool emmeans results across imputations (matching main effects approach)
pool_emmeans_subgroup_effects <- function(models, n_imputations) {

  # Set emmeans options to handle larger reference grids (matching main effects)
  emm_options(rg.limit = 130000)

  # Store results from each imputation
  all_emmeans <- list()
  all_contrasts <- list()
  all_risk_diffs <- list()

@@

    # Get marginal means on both log odds and response scales
    emm_logit <- emmeans(models[[i]], 
                         ~ ever_lapse_binary | dr_severity * any_treatment,
                         weights = "proportional")

    emm_response <- emmeans(models[[i]], 
                           ~ ever_lapse_binary | dr_severity * any_treatment,
                           weights = "proportional",
                           type = "response")

    # Get contrasts on log odds scale (for odds ratios)
    lapse_contrasts <- contrast(emm_logit, method = "revpairwise", 
                               by = c("dr_severity", "any_treatment"))

    # Get pairwise comparisons on response scale (for risk differences)
    risk_diffs <- pairs(emm_response, reverse = TRUE,
                       by = c("dr_severity", "any_treatment"))

    all_emmeans[[i]] <- as.data.frame(emm_response)
    all_contrasts[[i]] <- as.data.frame(lapse_contrasts)
    all_risk_diffs[[i]] <- as.data.frame(risk_diffs)
  }

  # Get structure from first imputation
  first_contrast <- all_contrasts[[1]]
  first_rd <- all_risk_diffs[[1]]
  n_subgroups <- nrow(first_contrast)

  # Initialize pooled results data frames
  pooled_or_results <- data.frame(
    dr_severity = first_contrast$dr_severity,
    any_treatment = first_contrast$any_treatment,
    contrast = first_contrast$contrast,
    log_odds = numeric(n_subgroups),
    SE_log_odds = numeric(n_subgroups),
    df = numeric(n_subgroups),
    stringsAsFactors = FALSE
  )

  pooled_rd_results <- data.frame(
    dr_severity = first_rd$dr_severity,
    any_treatment = first_rd$any_treatment,
    contrast = first_rd$contrast,
    risk_diff = numeric(n_subgroups),
    SE_risk_diff = numeric(n_subgroups),
    df = numeric(n_subgroups),
    stringsAsFactors = FALSE
  )

  # Pool each subgroup using Rubin's rules
  for (j in 1:n_subgroups) {
    # Pool odds ratio contrasts (log scale)
    log_odds_estimates <- sapply(all_contrasts, function(x) x$estimate[j])
    log_odds_SEs <- sapply(all_contrasts, function(x) x$SE[j])

    # Apply Rubin's rules for log odds
    Q_bar_log <- mean(log_odds_estimates)
    U_bar_log <- mean(log_odds_SEs^2)
    B_log <- var(log_odds_estimates)
    T_log <- U_bar_log + (1 + 1/n_imputations) * B_log

    # Degrees of freedom (Barnard-Rubin)
    lambda_log <- (B_log + B_log/n_imputations) / T_log
    df_log <- (n_imputations - 1) / (lambda_log^2)

    pooled_or_results$log_odds[j] <- Q_bar_log
    pooled_or_results$SE_log_odds[j] <- sqrt(T_log)
    pooled_or_results$df[j] <- df_log

    # Pool risk differences (probability scale)
    rd_estimates <- sapply(all_risk_diffs, function(x) x$estimate[j])
    rd_SEs <- sapply(all_risk_diffs, function(x) x$SE[j])

    # Apply Rubin's rules for risk differences
    Q_bar_rd <- mean(rd_estimates)
    U_bar_rd <- mean(rd_SEs^2)
    B_rd <- var(rd_estimates)
    T_rd <- U_bar_rd + (1 + 1/n_imputations) * B_rd

    # Degrees of freedom
    lambda_rd <- (B_rd + B_rd/n_imputations) / T_rd
    df_rd <- (n_imputations - 1) / (lambda_rd^2)

    pooled_rd_results$risk_diff[j] <- Q_bar_rd
    pooled_rd_results$SE_risk_diff[j] <- sqrt(T_rd)
    pooled_rd_results$df[j] <- df_rd
  }

  return(list(
    odds_ratios = pooled_or_results,
    risk_differences = pooled_rd_results,
    emmeans_list = all_emmeans
  ))
}

# Pool contrasts across all imputations
pooled_results <- pool_emmeans_subgroup_effects(models, n_imputations)

# Process odds ratio results
or_effects <- pooled_results$odds_ratios |>
  mutate(
    t_stat = log_odds / SE_log_odds,
    p.value = 2 * pt(-abs(t_stat), df = df),
    CI_Lower_log = log_odds - qt(0.975, df = df) * SE_log_odds,
    CI_Upper_log = log_odds + qt(0.975, df = df) * SE_log_odds,
    OR = exp(log_odds),
    OR_CI_Lower = exp(CI_Lower_log),
    OR_CI_Upper = exp(CI_Upper_log)
  )

# Process risk difference results
rd_effects <- pooled_results$risk_differences |>
  mutate(
    t_stat = risk_diff / SE_risk_diff,
    p.value = 2 * pt(-abs(t_stat), df = df),
    CI_Lower = risk_diff - qt(0.975, df = df) * SE_risk_diff,
    CI_Upper = risk_diff + qt(0.975, df = df) * SE_risk_diff
  )

# Create odds ratio table
or_table <- data.frame(
  DR_Severity = or_effects$dr_severity,
  Treatment = or_effects$any_treatment,
  OR = round(or_effects$OR, 3),
  `OR_95CI` = paste0("(", round(or_effects$OR_CI_Lower, 3), ", ", 
                     round(or_effects$OR_CI_Upper, 3), ")"),
  P_Value = format.pval(or_effects$p.value, digits = 4, eps = 0.0001)
)

# Create risk difference table
rd_table <- data.frame(
  DR_Severity = rd_effects$dr_severity,
  Treatment = rd_effects$any_treatment,
  Risk_Diff = round(rd_effects$risk_diff, 4),
  `RD_95CI` = paste0("(", round(rd_effects$CI_Lower, 4), ", ", 
                     round(rd_effects$CI_Upper, 4), ")"),
  P_Value = format.pval(rd_effects$p.value, digits = 4, eps = 0.0001)
)

kable(or_table,
      caption = "Odds Ratios for Lapsing Effect Within Each Subgroup",
      col.names = c("DR Severity", "Treatment", "Odds Ratio", "95% CI", "P-value"),
      row.names = FALSE)

kable(rd_table,
      caption = "Risk Differences for Lapsing Effect Within Each Subgroup",
      col.names = c("DR Severity", "Treatment", "Risk Diff", "95% CI", "P-value"),
      row.names = FALSE)

# Store for visualization (use OR results for compatibility)
subgroup_effects_stored <- data.frame(
  DR_Severity = or_effects$dr_severity,
  Treatment = or_effects$any_treatment,
  Log_Odds = or_effects$log_odds,
  SE = or_effects$SE_log_odds,
  CI_Lower = or_effects$CI_Lower_log,
  CI_Upper = or_effects$CI_Upper_log,
  OR = or_effects$OR,
  OR_CI_Lower = or_effects$OR_CI_Lower,
  OR_CI_Upper = or_effects$OR_CI_Upper,
  P_Value = or_effects$p.value
)
```

### Understanding the Different Effect Measures

::: {.callout-note}
## Why are the confidence intervals so different between measures?

The three effect measures presented above capture different aspects of the lapsing effect:

1. **Odds Ratios (OR)**: 
   - Multiplicative scale comparing odds of outcome
   - Wider CIs because they incorporate uncertainty from multiple interaction terms
   - Non-collapsible measure that depends on covariate distribution

2. **Risk Differences (RD)**:
   - Additive scale showing absolute change in probability
   - Narrower CIs because probabilities are bounded [0,1]
   - More intuitive interpretation for clinical decision-making

3. **Average Marginal Effects (AME)** (shown later):
   - Average of individual-level effects across actual patients
   - Smallest CIs because averaging reduces variance
   - Accounts for actual covariate distribution in each subgroup

The OR confidence intervals are wider because:
- They must account for uncertainty in the main effect plus all relevant interaction terms
- The log scale amplifies uncertainty when exponentiating
- Small subgroups have fewer effective degrees of freedom
:::

@@

## Convergence Diagnostics

```{r convergence-diagnostics}
# Extract estimates across imputations for key subgroups
convergence_data <- data.frame()

for (i in 1:n_imputations) {
  # Get log odds estimates from each imputation for each subgroup
  contrast_df <- pooled_results$odds_ratios

  # Track No_DR + No_Treatment as reference
  ref_estimates <- sapply(1:n_imputations, function(imp) {
    df <- as.data.frame(pooled_results$emmeans_list[[imp]])
    subset_df <- df[df$dr_severity == "No_DR" & 
                   df$any_treatment == "No_Treatment" & 
                   df$ever_lapse_binary == "Lapsed", ]
    if (nrow(subset_df) > 0) {
      return(qlogis(subset_df$prob[1]))  # Convert prob to log odds
    } else {
      return(NA)
    }
  })

  # Track PDR + Any_Treatment as most complex interaction
  complex_estimates <- sapply(1:n_imputations, function(imp) {
    df <- as.data.frame(pooled_results$emmeans_list[[imp]])
    subset_df <- df[df$dr_severity == "PDR" & 
                   df$any_treatment == "Any_Treatment" & 
                   df$ever_lapse_binary == "Lapsed", ]
    if (nrow(subset_df) > 0) {
      return(qlogis(subset_df$prob[1]))
    } else {
      return(NA)
    }
  })
}

# Calculate running averages
running_avg_ref <- cumsum(ref_estimates) / seq_along(ref_estimates)
running_avg_complex <- cumsum(complex_estimates) / seq_along(complex_estimates)

# Create convergence plot
convergence_df <- data.frame(
  Imputation = rep(1:n_imputations, 2),
  Estimate = c(ref_estimates, complex_estimates),
  Running_Average = c(running_avg_ref, running_avg_complex),
  Subgroup = factor(rep(c("No_DR + No_Treatment", "PDR + Any_Treatment"), 
                        each = n_imputations))
)

# Plot convergence
library(ggplot2)
convergence_plot <- ggplot(convergence_df, aes(x = Imputation)) +
  geom_point(aes(y = Estimate, color = Subgroup), alpha = 0.5) +
  geom_line(aes(y = Running_Average, color = Subgroup), size = 1.2) +
  labs(title = "Convergence of Subgroup Effect Estimates",
       subtitle = "Points show individual estimates, lines show running averages",
       x = "Imputation Number",
       y = "Log Odds of Outcome (Lapsed)",
       color = "Subgroup") +
  theme_minimal() +
  theme(legend.position = "bottom")

print(convergence_plot)

# Summary statistics
convergence_stats <- data.frame(
  Subgroup = c("No_DR + No_Treatment", "PDR + Any_Treatment"),
  Mean = c(mean(ref_estimates), mean(complex_estimates)),
  SD = c(sd(ref_estimates), sd(complex_estimates)),
  CV = c(sd(ref_estimates)/abs(mean(ref_estimates)) * 100,
         sd(complex_estimates)/abs(mean(complex_estimates)) * 100),
  Final_Running_Avg = c(tail(running_avg_ref, 1), tail(running_avg_complex, 1))
) |>
  mutate(across(where(is.numeric), ~round(., 4)))

kable(convergence_stats,
      caption = "Convergence Statistics for Key Subgroups",
      col.names = c("Subgroup", "Mean", "SD", "CV (%)", "Final Running Avg"))
```
